---
title: "Ghana Rainfall Data Extraction and Analysis Workflow"
author: "Hanna"
date: "2025-03-19"
output: html_document
---



```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Introduction

This document provides a step-by-step workflow for extracting daily rainfall data from Google Earth Engine (GEE), processing it in R, and conducting exploratory data analysis. The data covers Ghana for the years 2023 and 2024, using the CHIRPS Daily Precipitation dataset.

# Extracting Data with Google Earth Engine

To extract the data, use the following Google Earth Engine (GEE) script in JavaScript:

```javascript
// Define Ghana boundary using the USDOS LSIB dataset
var ghana = ee.FeatureCollection("USDOS/LSIB_SIMPLE/2017")
  .filter(ee.Filter.eq("country_na", "Ghana"));

// Define time periods
var startDate2023 = '2023-01-01';
var endDate2023 = '2023-12-31';
var startDate2024 = '2024-01-01';
var endDate2024 = '2024-12-31';

// Load CHIRPS Daily Precipitation dataset for both years
var dataset2023 = ee.ImageCollection('UCSB-CHG/CHIRPS/DAILY')
  .filterDate(startDate2023, endDate2023)
  .select('precipitation')
  .map(function(img) { return img.unmask(0); }); // Fill missing values

var dataset2024 = ee.ImageCollection('UCSB-CHG/CHIRPS/DAILY')
  .filterDate(startDate2024, endDate2024)
  .select('precipitation')
  .map(function(img) { return img.unmask(0); });

// Function to extract mean rainfall over Ghana for each day
var extractRainfall = function(image) {
  var meanRainfall = image.reduceRegion({
    reducer: ee.Reducer.mean(),
    geometry: ghana.geometry(),  // Ensure correct boundary usage
    scale: 5000,
    maxPixels: 1e13
  }).get('precipitation');

  return ee.Feature(null, {
    'date': image.date().format('YYYY-MM-dd'),
    'rainfall': meanRainfall
  });
};

// Apply function over image collections
var results2023 = dataset2023.map(extractRainfall);
var results2024 = dataset2024.map(extractRainfall);

// Merge results
var allResults = results2023.merge(results2024);

// Export results to Google Drive
Export.table.toDrive({
  collection: allResults,
  description: 'Rainfall_Ghana_All_2023_2024',
  fileFormat: 'CSV'
});

// Display first image for validation
Map.centerObject(ghana, 6);
Map.addLayer(dataset2023.first(), {min: 0, max: 20, palette: ['blue', 'yellow', 'red']}, 'Rainfall 2023 Sample');
```

# Loading and Processing Data in R

Once the data is extracted from GEE, follow these steps to process and analyze it in R.

## Load Required Libraries

```{r setup, message=FALSE, warning=FALSE}
library(readr)
library(dplyr)
library(ggplot2)
```

## Load and Inspect Data

```{r load_data}
# Load rainfall dataset
rainfall_data <- read_csv("Rainfall_Ghana_All_2023_2024.csv")

#structure of dataset
str(rainfall_data)
head(rainfall_data)
dim(rainfall_data)
```

```stata
import delimited "Rainfall_Ghana_All_2023_2024.csv", clear
describe
list in 1/5
count
```

## Summary Statistics

```{r summary_stats}
summary(rainfall_data$rainfall)
```

```stata
summarize rainfall
```


## Visualizing Rainfall Trends

```{r rainfall_trend, fig.width=8, fig.height=5}
# Convert date column to Date format
rainfall_data <- rainfall_data %>%
  mutate(date = as.Date(date))

# Plot rainfall trend
ggplot(rainfall_data, aes(x = date, y = rainfall)) +
  geom_line(color = "blue") +
  labs(title = "Daily Rainfall in Ghana (2023-2024)", x = "Date", y = "Rainfall (mm)")
```


```stata
gen date2 = date(date, "YMD")
format date2 %td
drop date
rename date2 date
```


```{r cars}
summary(rainfall_data)
head(rainfall_data)
```

## Identifying Consecutive Dry Days (CDD)

```{r cdd_calculation}
rainfall_data <- rainfall_data %>%
  arrange(date) %>% #sorts the data by date
  mutate(dry_day = ifelse(rainfall < 1, 1, 0), 
         CDD = sequence(rle(dry_day)$lengths) * dry_day)
#creates a dry_day column:1 if rainfall < 1mm (dry day), 0 if rainfall ≥ 1mm (wet day).
#counts consecutive dry days (CDD): rle() (Run-Length Encoding) to detect consecutive dry days.

# View distribution of CDD
summary(rainfall_data$CDD)
```


```stata
twoway (line rainfall date), ///
    title("Daily Rainfall in Ghana (2023-2024)") ///
    xlabel(, angle(45)) ///
    ylabel(, grid) ///
    xtitle("Date") ytitle("Rainfall (mm)")
```



```{r rainfall anomaly index (RAI)}
rainfall_data <- rainfall_data %>%
  mutate(rai = (rainfall - mean(rainfall, na.rm = TRUE)) / sd(rainfall, na.rm = TRUE))
```

```{r standardized precipitation index (SPI)}
#install.packages("SPEI")
#install.packages("lubridate")
library(SPEI)
library(dplyr)
library(lubridate)

# Aggregate daily rainfall to monthly totals
rainfall_monthly <- rainfall_data %>%
  mutate(year_month = floor_date(date, "month")) %>%
  group_by(year_month) %>%
  summarise(monthly_rainfall = sum(rainfall, na.rm = TRUE))

# Compute 3-month SPEI
spei_3_month <- spei(rainfall_monthly$monthly_rainfall, scale = 3)

# Add SPEI values to dataset
rainfall_monthly$SPEI <- spei_3_month$fitted

# View results
head(rainfall_monthly)
```

```stata
gen dry_day = (rainfall < 1)

gen CDD = .
replace CDD = 1 if dry_day == 1 & dry_day[_n-1] == 0
replace CDD = CDD[_n-1] + 1 if dry_day == 1 & dry_day[_n-1] > 0
replace CDD = 0 if dry_day == 0

summarize CDD
```

```stata
histogram CDD, ///
    title("Distribution of Consecutive Dry Days (CDD)") ///
    xlabel(0(5)30) ///
    ylabel(, grid) ///
    xtitle("Consecutive Dry Days") ytitle("Frequency") ///
    color(blue)
```

```{r save processed data}
#install.packages("readr")
library(readr)
write_csv(rainfall_data, "Processed_Rainfall_Ghana_2023_2024.csv")
head(rainfall_data)
```

